\name: Deterministic and Coherent Implicits

### 2018-06-28 jfp/src/Translation.tex
real update theorem statements

\begin{theorem}[Type-Preserving Translation]\label{thm:type:preservation} Let $e$ be a $\name$
  expression, $\rulet$ be a type, $\tenv$ be a type environment and $E$ be a System F expression. If
  $\wte{\tenv}{e}{\rulet}{E}$, then $\fwte{\etrans{\tenv}}{E}{\ttrans[\rulet]}$.
\end{theorem}
Here we define the translation of the type environment form $\name$ to System F as:
\begin{equation*}
\begin{array}{rcl@{\hspace{2cm}}rcl}
\etrans{\epsilon} & = & \epsilon & \etrans{\tenv,\alpha} & = & \etrans{\tenv}, \alpha \\
\etrans{\tenv,x : \rulet} & = & \etrans{\tenv}, x : \ttrans[\rulet] &
\etrans{\tenv, \envi{\rulet}{x}} & = & \etrans{\tenv}, x : \ttrans[\rulet]
\end{array}
\end{equation*}

@@

Moreover, we can express three key properties of Figure~\ref{fig:resolution2}'s
definition of resolution in terms of the generated evidence.
Firstly, the deterministic version of resolution is a sound variation on the original ambiguous resolution.
\begin{theorem}[Soundness]
Figure~\ref{fig:resolution2}'s definition of resolution is sound (but
incomplete) with respect to Figure~\ref{fig:resolution1}'s definition.
\[\forall \tenv, \rulet, E: \quad\quad \dres{\tenv}{\rulet}{E} \quad\Rightarrow\quad \ares{\tenv}{\rulet}{E} \]
\end{theorem}
Secondly, the deterministic resolution is effectively deterministic:
\begin{theorem}[Determinacy]
The generated evidence of resolution is uniquely determined.
\[\forall \tenv, \rulet, E_1, E_2: \quad\quad 
     \dres{\tenv}{\rulet}{E_1} ~\wedge~ 
     \dres{\tenv}{\rulet}{E_2} \quad\Rightarrow\quad E_1 = E_2 \]
\end{theorem}
It follows immediately that deterministic resolution is also coherent,
which for our elaboration-based setting is formulated, following Biernacki and Polesiuk~\shortcite{dariusz}, in
terms of \emph{contextual equivalence}~\cite{morris}.
\begin{corollary}[Coherence]
\[\forall \tenv, \rulet, E_1, E_2: \quad\quad 
     \dres{\tenv}{\rulet}{E_1} ~\wedge~ 
     \dres{\tenv}{\rulet}{E_2} \quad\Rightarrow\quad 
     \fctx{\etrans{\tenv}}{E_1}{E_2}{\ttrans[\rulet]} \]
\end{corollary}
Here the contextual equivalence judgment $\fctx{\tenv}{E_1}{E_2}{T}$ captures
that $E_1$ and $E_2$ behave the same in any well-typed program context. We do
not give the formal definition here\footnote{A definition can be found at \url{https://www.seas.harvard.edu/courses/cs252/2017fa/lectures/lec02-contextual-equiv.pdf}} as it is not required to prove the
corollary; we only need to know that it is an equivalence relation, and, more
spefically, that it is reflexive, i.e., any well-typed System~F term is
contextually equivalent to itself. 
% \bruno{Should we not mention that contextual equivalence for System F
%   is well-studied, thus interested readers can consult the definition
%   in other works?}

Thirdly, on top of the immediate coherence of deterministic resolution, 
an additional stability property holds.  
\begin{lemma}[Stability of Resolution]
Resolution is stable under type substitution.
\[\forall \tenv,\alpha,\tenv',\sigma,\rulet, E: \enskip\
\dres{\tenv,\alpha,\tenv'}{\rulet}{E} \enskip\wedge\enskip \tenv \vdash \sigma
\enskip\Rightarrow\enskip 
\dres{\tenv,\tenv'[\sigma/\alpha]}{\rulet[\sigma/\alpha]}{E[\ttrans[\sigma]/\alpha]} \]
\end{lemma}%
This is a key lemma to establish that static reduction of type application 
preserves typing and elaboration.
\begin{lemma}[Stability of Typing under Type Application Reduction]
Static reduction of type application preserves typing.
\[\forall \tenv,\alpha,\sigma,\rulet, E: \enskip\
\wte{\tenv}{(\Lambda \alpha.e)\,\suty}{\rulet}{(\Lambda \alpha.E)\,\ttrans[\suty]}
\enskip\Rightarrow\enskip 
\wte{\tenv}{e[\suty/\alpha]}{\rulet}{E[\ttrans[\suty]/\alpha]}
   \]
\end{lemma}%
This, together with another property of System~F's contextual equivalence:
\[ \forall \tenv, \alpha, E, T, T': \enskip\
     \fwte{\tenv}{(\Lambda \alpha.E)\,T}{T'} \enskip\Rightarrow\enskip 
     \fctx{\tenv}{(\Lambda \alpha.E)\,T}{E[T/\alpha]}{T'}\]
allows us to conclude the correctness of static reduction of type application.
\begin{theorem}[Correctness of Type Application Reduction]
If a type application and its reduced form elaborate to two System~F terms,
those terms are contextually equivalent.
\begin{equation*}
\forall \tenv, \alpha, e, \suty, E1, E2 : \enskip\\
\wte{\tenv}{(\Lambda \alpha.e)\,\suty}{\rulet}{E_1} ~\wedge~
\wte{\tenv}{e[\suty/\alpha]}{\rulet}{E_2}
\enskip\Rightarrow\enskip 
\fctx{\etrans{\tenv}}{E_1}{E_2}{\ttrans[\rulet]}
\end{equation*}
\end{theorem}

@@

\begin{theorem}[Partial Correctness]
Let $\tenv$ be a type environment, $\rulet$ be a type and E be a System F expression.
Assume that $\unambig{\epsilon}{\rulet}$ and also $\forall \rulet_i \in \tenv: \unambig{\epsilon}{\rulet_i}$.
Then $\dres{\tenv}{\rulet}{E}$ if and only if $\adres{\tenv}{\rulet}{E}$.
\end{theorem}

>>>>

\begin{theorem}[Type-Preserving Translation]\label{thm:type:preservation} Let $e$ be a $\name$
  expression, $\rulet$ be a type, $\tenv$ a type environment and $E$ a System F expression. If
  $\wte{\tenv}{e}{\rulet}{E}$, then $\fwte{\etrans{\tenv}}{E}{\ttrans[\rulet]}$.
\end{theorem}
Here we define the translation of the type environment form $\name$ to System F as:
\begin{equation*}
\begin{array}{rcl@{\hspace{2cm}}rcl}
\etrans{\epsilon} & = & \epsilon & \etrans{\tenv,\alpha} & = & \etrans{\tenv}, \alpha \\
\etrans{\tenv,x : \rulet} & = & \etrans{\tenv}, x : \ttrans[\rulet] &
\etrans{\tenv, \envi{\rulet}{x}} & = & \etrans{\tenv}, x : \ttrans[\rulet]
\end{array}
\end{equation*}

@@

Moreover, we can express three key properties of Figure~\ref{fig:resolution2}'s
definition of resolution in terms of the generated evidence.
Firstly, the deterministic version of resolution is a sound variation on the original ambiguous resolution.
\begin{theorem}[Soundness]
Figure~\ref{fig:resolution2}'s definition of resolution is sound (but
incomplete) with respect to Figure~\ref{fig:resolution1}'s definition.
\[\forall \tenv, \rulet, E: \quad\quad \dres{\tenv}{\rulet}{E} \quad\Rightarrow\quad \ares{\tenv}{\rulet}{E} \]
\end{theorem}
Secondly, the deterministic resolution is effectively deterministic:
\begin{theorem}[Determinacy]
%The generated evidence of resolution is uniquely determined.
Let $\tenv$ be a type environment, $\rulet$ be a type and $E_1$, $E_2$ be system F expressions.
Assume that $\unambig{}{\rulet}$ and $\forall ?\rulet_i\in\tenv$, $\unambig{}{\rulet_i}$. Then
\[ %\forall \tenv, \rulet, E_1, E_2: \quad\quad 
     \dres{\tenv}{\rulet}{E_1} ~\wedge~ 
     \dres{\tenv}{\rulet}{E_2} \quad\Rightarrow\quad E_1 = E_2 \]
\end{theorem}
It follows immediately that deterministic resolution is also coherent,
which for our elaboration-based setting is formulated, following Biernacki and Polesiuk~\shortcite{dariusz}, in
terms of \emph{contextual equivalence}~\cite{morris}.
\begin{corollary}[Coherence]
Let $\tenv$ be a type environment, $\rulet$ be a type and $E_1$, $E_2$ be system F expressions.
Assume that $\unambig{}{\rulet}$ and $\forall ?\rulet_i\in\tenv$, $\unambig{}{\rulet_i}$. Then
\[ %\forall \tenv, \rulet, E_1, E_2: \quad\quad 
     \dres{\tenv}{\rulet}{E_1} ~\wedge~ 
     \dres{\tenv}{\rulet}{E_2} \quad\Rightarrow\quad 
     \fctx{\etrans{\tenv}}{E_1}{E_2}{\ttrans[\rulet]} \]
\end{corollary}
Here the contextual equivalence judgment $\fctx{\tenv}{E_1}{E_2}{T}$ captures
that $E_1$ and $E_2$ behave the same in any well-typed program context. We do
not give the formal definition here\footnote{A definition can be found at \url{https://www.seas.harvard.edu/courses/cs252/2017fa/lectures/lec02-contextual-equiv.pdf}} as it is not required to prove the
corollary; we only need to know that it is an equivalence relation, and, more
spefically, that it is reflexive, i.e., any well-typed System~F term is
contextually equivalent to itself. 
% \bruno{Should we not mention that contextual equivalence for System F
%   is well-studied, thus interested readers can consult the definition
%   in other works?}

Thirdly, on top of the immediate coherence of deterministic resolution, 
an additional stability property holds.  
\begin{lemma}[Stability of Resolution]
Resolution is stable under type substitution.
\[\forall \tenv,\alpha,\tenv',\suty,\rulet, E: \enskip\
\dres{\tenv,\alpha,\tenv'}{\rulet}{E} \enskip\wedge\enskip \tenv \vdash \sigma
\enskip\Rightarrow\enskip 
\dres{\tenv,\tenv'[\sigma/\alpha]}{\rulet[\sigma/\alpha]}{E[\ttrans[\sigma]/\alpha]} \]
\end{lemma}%
This is a key lemma to establish that static reduction of type application 
preserves typing and elaboration.
\begin{lemma}[Stability of Typing under Type Application Reduction]
Static reduction of type application preserves typing.
\[\forall \tenv,\alpha,\sigma,\rulet, E: \enskip\
\wte{\tenv}{(\Lambda \alpha.e)\,\suty}{\rulet}{(\Lambda \alpha.E)\,\ttrans[\suty]}
\enskip\Rightarrow\enskip 
\wte{\tenv}{e[\suty/\alpha]}{\rulet}{E[\ttrans[\suty]/\alpha]}
   \]
\end{lemma}%
This, together with another property of System~F's contextual equivalence:
\[ \forall \tenv, \alpha, E, T, T': \enskip\
     \fwte{\tenv}{(\Lambda \alpha.E)\,T}{T'} \enskip\Rightarrow\enskip 
     \fctx{\tenv}{(\Lambda \alpha.E)\,T}{E[T/\alpha]}{T'}\]
allows us to conclude the correctness of static reduction of type application.
\begin{theorem}[Correctness of Type Application Reduction]
If a type application and its reduced form elaborate to two System~F terms,
those terms are contextually equivalent.
\begin{equation*}
\forall \tenv, \alpha, e, \suty, E1, E2 : \enskip\\
\wte{\tenv}{(\Lambda \alpha.e)\,\suty}{\rulet}{E_1} ~\wedge~
\wte{\tenv}{e[\suty/\alpha]}{\rulet}{E_2}
\enskip\Rightarrow\enskip 
\fctx{\etrans{\tenv}}{E_1}{E_2}{\ttrans[\rulet]}
\end{equation*}
\end{theorem}

@@

\begin{theorem}[Partial Correctness]
Let $\tenv$ be a type environment, $\rulet$ be a type and E be a System F expression.
Assume that $\unambig{}{\rulet}$ and also $\forall ?\rulet_i \in \tenv: \unambig{}{\rulet_i}$.
Then $\dres{\tenv}{\rulet}{E}$ if and only if $\adres{\tenv}{\rulet}{E}$.
\end{theorem}
